---
title: "DSA311_proj"
output: html_document
date: "2024-03-16"
---

#read the dataset and split into test/train set 
```{r}
resale <- read.csv("final_hdb_resale_2023.csv", stringsAsFactors = TRUE)
variables_to_keep <- c("resale_price",
  "month", "area", "lat", "long", "floor_area_sqm", "flat_model", 
  "remaining_lease", "avg_storey", "no_of_rooms", "closestMall", 
  "timeToMall", "mallsIn10km", "closestBeach", "timeToBeach", 
  "beachesIn10km", "closestMRTStation", "timeToMRTStation", 
  "MRTStationsIn10km", "closestPrimarySch", "timeToPrimarySch", 
  "PrimarySchsIn10km", "below1000Pop", "below1500Pop", "below2000Pop", 
  "below2500Pop", "below3000Pop", "below4000Pop", "below5000Pop", 
  "below6000Pop", "below7000Pop", "below8000Pop", "below9000Pop", 
  "below10000Pop", "below11000Pop", "below12000Pop", "above12000Pop"
)

resale <- subset(resale, select = variables_to_keep)


set.seed(444)
train <- sample(1:nrow(resale),8000)
test <- -train
```

############
1. EDA 
############

1.1 Price Distribution By Area 
```{r}
library(ggplot2)
ggplot(resale, aes(x = area, y = resale_price)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Price Distribution by Area", x = "Area", y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

1.2 Distribution of flat type(based on number of rooms) across different areas 
```{r}
#Distribution of flat type across different areas 
ggplot(resale, aes(x = area, fill = factor(no_of_rooms))) +
  geom_bar() +
  labs(title = "Distribution of Flat Types Across Different Areas",
       x = "Area", y = "Count") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")
```

1.3 Scatter plot of time to amenities vs resale price 
```{r}
library(gridExtra)
plot1 <- ggplot(resale, aes(x = timeToMall, y = resale_price)) +
  geom_point() +
  labs(title = "Time to Mall vs. Resale Price")+ 
  geom_smooth(method = "lm", se = FALSE)

plot2 <-ggplot(resale, aes(x = timeToBeach, y = resale_price)) +
  geom_point() +
  labs(title = "Time to Beach vs. Resale Price") + 
  geom_smooth(method = "lm", se = FALSE)

plot3 <- ggplot(resale, aes(x = timeToMRTStation, y = resale_price)) +
  geom_point() +
  labs(title = "Time to MRT Station vs. Resale Price") + 
  geom_smooth(method = "lm", se = FALSE)

final_plot <- grid.arrange(plot1, plot2, plot3, ncol = 3)

# Display the final plot
print(final_plot)
```

1.4 Correlation plot of numerical variables 
```{r}
#install.packages("corrplot")
library(corrplot)

#remove categorical data 
correlation1 <- resale |>
  select_if(~ !is.factor(.)) 

cor_matrix <- cor(correlation1)
corrplot(cor_matrix, method = "circle",tl.col = "black", tl.srt = 45)
```
```{r}

library(corrplot)

#remove categorical data 
correlation1 <- resale |>
  select_if(~ !is.factor(.)) 

columns <- c(
   "lat", "long", "floor_area_sqm", 
  "remaining_lease", "avg_storey", "no_of_rooms", 
  "timeToMall", "mallsIn10km", "timeToBeach", 
  "beachesIn10km", "timeToMRTStation", 
  "MRTStationsIn10km", "timeToPrimarySch", 
  "PrimarySchsIn10km")

correlation1 <- subset(correlation1, select = columns)

cor_matrix <- cor(correlation1)
corrplot(cor_matrix, method = "circle",tl.col = "black", tl.srt = 45)
```

############
2. Clustering Analysis 
############

2.1 Data preprocessing for clustering 
```{r}
clustering_data <- resale
#convert month values to integer from 1 -12 - Numeric Encoding 
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
clustering_data$month <- match(resale$month, months)

#do one hot encoding for all categorial variables - to be used in clustering 
cols_to_encode <- clustering_data |>
  select_if(~ is.factor(.)) |>
  names()

clustering_data[,cols_to_encode] <- sapply(clustering_data[, cols_to_encode], function(x) as.integer(x))

#standardize the columns for clustering
clustering_data <- as.data.frame(scale(clustering_data))
#head(clustering_data)

```
2.2 Kmeans clustering 
```{r}
library(stats)
library(cluster)
#range of k values to try
k_values <- 1:15
wcss <- numeric(length(k_values))
silhouette_scores <- numeric(length(k_values))

for (k in k_values) {
  kmeans_result <- kmeans(clustering_data, centers = k)
  wcss[k] <- kmeans_result$tot.withinss
  ssi <- summary(silhouette(kmeans_result$cluster, dist(clustering_data)))
  silhouette_scores[k] <- ssi[4][[1]]
}

# Plot the elbow curve
plot(k_values, wcss, type = "b", pch = 19, frame = FALSE, 
     xlab = "Number of clusters (k)", ylab = "Total within-cluster sum of squares (WCSS)",
     main = "Elbow Method for Optimal k")

#Plot the silhouette score 
plot(k_values, silhouette_scores, type = "b", 
     xlab = "Number of Clusters (k)", ylab = "Silhouette Score",
     main = "Silhouette Score vs. Number of Clusters")

#clusters are not well separated based on silhouette score and wcss
silhouette_scores[which.max(silhouette_scores)]
wcss[which.min(wcss)]
```

2.3 Hierarchical clustering
```{r}
library(cluster)
h_values <- 1:15

silhouette_scores <- numeric(length(h_values))

for (h in h_values){
  cl <- hclust(dist(clustering_data), method= "complete")
  hc_cluster <- cutree(cl, h = h)
  ssi <- summary(silhouette(hc_cluster, dist(clustering_data)))
  silhouette_scores[h]<- ssi[4][[1]]
}

plot(h_values, silhouette_scores, type = "b", 
     xlab = "Number of Clusters (h)", ylab = "Silhouette Score",
     main = "Silhouette Score vs. Number of Clusters (Hierarchical)")
silhouette_scores[which.max(silhouette_scores)] #0.3041983

```
############
Tree Based Models 
############

#Data pre processing for BART/Gradient boosting
```{r}
library(dplyr)
#Integer Encoding 
new_df <- resale
str(new_df)
# Do one hot encoding for all categorical variables - to be used in clustering  
cols_to_encode <- new_df |>   
  select_if(~ is.factor(.)) |>   
  names()  
integer_encoding_df <- new_df
integer_encoding_df[,cols_to_encode] <- sapply(integer_encoding_df[, cols_to_encode], function(x) as.integer(x))
#head(integer_encoding_df)
#Target Encoding 
# Do target encoding for all categorical variables - to be used in clustering
target_encode <- function(data, column, target) {
  # Check if column exists in data
  if (!(column %in% names(data))) {
    stop("Column not found in the data frame.")
  }
  # Check if target variable exists in data
  if (!(target %in% names(data))) {
    stop("Target variable not found in the data frame.")
  }
  # Check if column is a factor
  if (!is.factor(data[[column]])) {
    stop("Column is not a factor.")
  }
  # Check if column and target have the same length
  if (length(data[[column]]) != length(data[[target]])) {
    stop("Column and target variable must have the same length.")
  }
  # Calculate mean of target variable for each category in the specified column
  encoding <- tapply(data[[target]], data[[column]], mean)
  # Map the mean values to the corresponding categories and assign to a new column
  data[[paste0(column, "_encoded")]] <- encoding[data[[column]]]
  # Return the modified data frame with target-encoded column
  return(data)
} # function to do target encoding
cols_to_encode <- names(new_df)[sapply(new_df, is.factor)] # Select categorical columns for target encoding
for (col in cols_to_encode) {
  new_df <- target_encode(data=new_df, column=col,target="resale_price")
} # Perform target encoding
factor_cols <- sapply(new_df, is.factor)
target_encoding_df <- new_df[, !factor_cols]
#head(target_encoding_df)
```


#Bayesian Additive Regression Trees
```{r}
#install.packages("BART")
library(BART)

#Integer Encoding
X <- subset(integer_encoding_df, select = -resale_price)
y <- integer_encoding_df$resale_price
Xtrain <- X[train,]
ytrain <- y[train]
Xtest <- X[test,]
ytest <- y[test]

bartfit <- gbart(Xtrain,ytrain,x.test = Xtest) #use default setting
yhat.bart <- bartfit$yhat.test.mean
(mean((ytest - yhat.bart)^2))^0.5

# Target Encoding
X <- subset(target_encoding_df, select = -resale_price)
y <- target_encoding_df$resale_price
Xtrain <- X[train,]
ytrain <- y[train]
Xtest <- X[test,]
ytest <- y[test]

bartfit <- gbart(Xtrain,ytrain,x.test = Xtest) #use default setting
yhat.bart <- bartfit$yhat.test.mean
(mean((ytest - yhat.bart)^2))^0.5

#No encoding of categorical variable
X <- subset(resale, select = -resale_price)
y <- resale$resale_price
Xtrain <- X[train,]
ytrain <- y[train]
Xtest <- X[test,]
ytest <- y[test]

bartfit <- gbart(Xtrain,ytrain,x.test = Xtest) #use default setting
yhat.bart <- bartfit$yhat.test.mean
(mean((ytest - yhat.bart)^2))^0.5

```

#Gradient Boosting Regressor
```{r}
set.seed(444)
train <- sample(1:nrow(resale),8000)
test <- -train
#install.packages("gbm")
library(gbm)

boosting <- function(data, interaction.depth, n.trees) {
  boost.resale <- gbm(resale_price ~., data = data[train,], distribution = "gaussian", n.trees = n.trees, interaction.depth = interaction.depth)
  yhat.boost <- predict(boost.resale, newdata = data[test,], n.trees = n.trees) #default lambda = 0.001
  return((mean((yhat.boost - data[test,"resale_price"])^2))^0.5)
}

#No encoding of categorical variables
boosting(resale, 4, 1000)
boosting(resale, 4, 2000)
boosting(resale, 4, 3000)
boosting(resale, 2, 1000)
boosting(resale, 4, 2500)


#Integer Encoding
boosting(integer_encoding_df, 4, 1000)
boosting(integer_encoding_df, 4, 2000)
boosting(integer_encoding_df, 4, 3000)
boosting(integer_encoding_df, 2, 1000)
boosting(integer_encoding_df, 4, 2500)


#Target Encoding
boosting(target_encoding_df, 4, 1000)
boosting(target_encoding_df, 4, 2000)
boosting(target_encoding_df, 4, 3000)
boosting(target_encoding_df, 2, 1000)
boosting(target_encoding_df, 4, 2500) 



```

```{r}
#final model with all the data 
library(gbm)
boost.resale <- gbm(resale_price ~., data = target_encoding_df, distribution = "gaussian", n.trees = 2500, interaction.depth = 4)
summary(boost.resale)
yhat.boost <- predict(boost.resale, newdata = target_encoding_df[test,], n.trees = 2500) #default lambda = 0.001
(mean((yhat.boost - target_encoding_df[test,"resale_price"])^2))^0.5

```


```{r}
#predicting the price of a flat with the final model 

new_data = data.frame(month = "Jan", area = "Bukit Merah" , lat = 1.286492541, long = 103.8214336, floor_area_sqm = 83 , flat_model = "Model A", remaining_lease = 94, avg_storey= 20, no_of_rooms = 4 ,closestMall = "Tiong Bahru Plaza", timeToMall = 2.216666667, mallsIn10km = 69 , closestBeach = "Palawan Beach", timeToBeach = 14.16666667, beachesIn10km = 5, closestMRTStation = "Redhill", timeToMRTStation =9.733333333 , MRTStationsIn10km = 63, closestPrimarySch = "Alexandra Primary School", timeToPrimarySch = 1.75, PrimarySchsIn10km = 31 , below1000Pop =8.5 , below1500Pop = 9.2, below2000Pop = 6.9, below2500Pop = 6.6, below3000Pop = 4.4, below4000Pop = 8.3, below5000Pop = 7, below6000Pop = 5.5, below7000Pop = 4.7, below8000Pop = 4.2, below9000Pop = 3, below10000Pop = 2.1, below11000Pop = 2.2, below12000Pop = 1.5, above12000Pop = 8)

#target encoding for the categorical variables - month, area, flat_model, closestMall, closestBeach, closestMRTStation

library(dplyr)
calculate_mean_price <- function(data, select_col, target_var, value_of_interest) {
  result <- data |>
    group_by({{target_var}}) |>
    summarise(mean_price = mean({{select_col}}))
  
  mean_price <- result[result[[as.character(substitute(target_var))]] == value_of_interest, "mean_price"]
  return(mean_price)
}

mean_price_jan <- calculate_mean_price(resale, resale_price, month, "Jan")
mean_price_area <- calculate_mean_price(resale, resale_price, area, "Bukit Merah")
mean_price_flat_model <- calculate_mean_price(resale, resale_price, flat_model, "Model A")
mean_price_mall <- calculate_mean_price(resale, resale_price, closestMall, "Tiong Bahru Plaza")
mean_price_beach <- calculate_mean_price(resale, resale_price, closestBeach, "Palawan Beach")
mean_price_mrt <- calculate_mean_price(resale, resale_price, closestMRTStation, "Redhill")
mean_price_sch <- calculate_mean_price(resale, resale_price, closestPrimarySch, "Alexandra Primary School")

#modify the new data 
new_data$month_encoded <- mean_price_jan$mean_price
new_data$area_encoded <- mean_price_area$mean_price
new_data$flat_model_encoded <- mean_price_flat_model$mean_price
new_data$closestMall_encoded <- mean_price_mall$mean_price
new_data$closestBeach_encoded <- mean_price_beach$mean_price
new_data$closestMRTStation_encoded <- mean_price_mrt$mean_price
new_data$closestPrimarySch_encoded <- mean_price_sch$mean_price

#prediction
predict(boost.resale, newdata = new_data, n.trees = 2500)

```


